A Linear Solution of the Linearity Challenge

### 2025-12-25 paper/Reduction.lagda.tex
global revision

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ occurs on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is found in the splitting
$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

Note that \AgdaInductiveConstructor{r-close} (and several other reduction
constructors) quantifies over $p_0$ and $q_0$ which both prove the splitting
$\Context \simeq \AgdaInductiveConstructor{[]} + \Context$. Since the left
context is empty, these splittings must be equal and made of a sequence of
\AgdaInductiveConstructor{>} applications followed by
\AgdaInductiveConstructor{•}. In general, Agda will not be able to ``see'' that
they are definitionally equal, hence it is easier to quantify them separately so
that we do not have to worry about proving their equality whenever we wish to
apply this reduction.

@@

The remaining constructors that describe the base reductions follow a similar
pattern, noting that the implementation of \RWeaken and \RContract makes use of
auxiliary functions that respectively weaken and contract the typing context of
the resulting process as shown in \Cref{tab:semantics}. It is worth glancing at
the implementation of \RExists since it involves a non-trivial rewriting of
types:

@@

Recalling the definitions of \AgdaInductiveConstructor{ex} and
\AgdaInductiveConstructor{all} from \Cref{sec:process-agda}, we note that $P$
has type
$\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
and $(F~B)$ is a process of type
$\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
In order for these two process to be composable in a cut, it must be the case
that $\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A) \equiv
\AgdaFunction{dual}~(\AgdaFunction{subst}~[~B~/]~A)$, which was proved in
\Cref{sec:type-agda} under the name \AgdaFunction{dual-subst}. Thanks to the
implicit rewriting rule, we do not have to rewrite the index in the type of
$(F~B)$, which is silently accepted as is.

>>>>

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ occurs on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is encoded in the splitting
$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

Note that \AgdaInductiveConstructor{r-close} (and several other reduction
constructors) quantifies over $p_0$ and $q_0$ which both prove the splitting
$\Context \simeq \AgdaInductiveConstructor{[]} + \Context$. Since the left
partition is empty, these splittings must be equal and made of a sequence of
\AgdaInductiveConstructor{>} applications followed by
\AgdaInductiveConstructor{•}. In general, Agda will not be able to ``see'' that
they are definitionally equal, hence it is easier to quantify them separately so
that we do not have to prove their equality whenever we wish to apply this
reduction.

@@

The remaining constructors that describe the base reductions follow a similar
pattern, except for the implementation of \RWeaken and \RContract which require
auxiliary functions to respectively weaken and contract the typing context of
the resulting process as shown in \Cref{tab:semantics}. It is worth glancing at
the implementation of \RExists since it involves a non-trivial rewriting of
types:

@@

Recalling the definitions of \AgdaInductiveConstructor{ex} and
\AgdaInductiveConstructor{all} from \Cref{sec:process-agda}, we note that $P$
has type
$\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
and $F~B$ is a process of type
$\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
In order for these two processes to be composable in a cut, it must be the case
that $\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A) \equiv
\AgdaFunction{dual}~(\AgdaFunction{subst}~[~B~/]~A)$, which was proved in
\Cref{sec:type-agda} under the name \AgdaFunction{dual-subst}. Thanks to the
implicit rewriting rule, we do not have to rewrite the index in the type of
$F~B$, which is silently accepted as is.

### 2025-12-25 paper/Safety.lagda.tex
global revision

Note the presence of two constructors \AgdaInductiveConstructor{cut-l} and
\AgdaInductiveConstructor{cut-r} which distinguish on whether the hole is found
in the left (respectively, right) sub-context, as per the grammar of reduction
contexts in \Cref{sec:properties}.

>>>>

Note the presence of two constructors \AgdaInductiveConstructor{cut-l} and
\AgdaInductiveConstructor{cut-r} which distinguish on whether the hole is found
in the left (respectively, right) sub-context, as per the grammar of reduction
contexts given in \Cref{sec:properties}.

### 2025-12-25 paper/Type.lagda.tex
global revision

The representation of types is standard, with one constructor for each of the
forms described in \Cref{sec:types}. We start by defining \emph{pre-types} as an
indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
type variables and we use de Bruijn indices for denoting type variables. In this
way, we make sure that pre-types are well-scoped.

@@

Note the constructors \AgdaInductiveConstructor{var} and
\AgdaInductiveConstructor{rav}, which respectively represent type variables and
their dual, and the quantifiers \AgdaInductiveConstructor{`∀} and
\AgdaInductiveConstructor{`∃} which increase the number of bound type variables
in the scoped (pre)type.

The dual of a (pre)type is computed by the following function:

@@

This property is so important in the rest of the formalisation that it is worth
defining an implicit rewriting rule so that Agda can autonomously apply it
whenever possible. This is achieved by means of the following pragma
directive.\footnote{The directive is effective provided that the option
\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or in
the module's preamble.}

@@

Here we are supposed to prove
$\AgdaFunction{subst}~\sigma~(\AgdaFunction{dual}~(\AgdaInductiveConstructor{rav}~x))
  \equiv
  \AgdaFunction{dual}~(\AgdaFunction{subst}~\sigma~(\AgdaInductiveConstructor{rav}~x))$
  which is definitionally the same of $\sigma~x \equiv
  \AgdaFunction{dual}~(\AgdaFunction{dual}~(\sigma~x))$.
%
We could easily prove this equivalence by invoking \AgdaFunction{dual-inv}, but
thanks to the rewriting rule that we have added just above a use of
\AgdaInductiveConstructor{refl} suffices. In this case the saved effort is
negligible, but in later results, where it is necessary to use
\AgdaFunction{dual-inv} for rewriting part of the \emph{index} of some type
families, having an implicit rewriting rule allows us to avoid writing some
quite obscure Agda code.

>>>>

The representation of types is standard, with one constructor for each of the
forms described in \Cref{sec:types}. We start by defining \emph{pre-types} as an
indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
type variables and we use de Bruijn indices for denoting type variables. In this
way, we make sure that pre-types are well scoped.

@@

Note the constructors \AgdaInductiveConstructor{var} and
\AgdaInductiveConstructor{rav}, which respectively represent type variables and
their dual, and the quantifiers \AgdaInductiveConstructor{`∀} and
\AgdaInductiveConstructor{`∃} which increase the number of bound type variables
in the scoped pre-type.

The dual of a pre-type is computed by the following function:

@@

This property is so important in the rest of the formalisation that it is worth
defining an implicit rewriting rule so that Agda can autonomously apply it
whenever possible. This is achieved by means of the following pragma
directive.\footnote{The directive is effective provided that the option
\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or
within a pragma in the module's source code.}

@@

Here we are supposed to prove
$\AgdaFunction{subst}~\sigma~(\AgdaFunction{dual}~(\AgdaInductiveConstructor{rav}~x))
  \equiv
  \AgdaFunction{dual}~(\AgdaFunction{subst}~\sigma~(\AgdaInductiveConstructor{rav}~x))$
  which is definitionally the same as $\sigma~x \equiv
  \AgdaFunction{dual}~(\AgdaFunction{dual}~(\sigma~x))$.
%
We could easily prove this equivalence by invoking \AgdaFunction{dual-inv}, but
thanks to the rewriting rule that we have added earlier a use of
\AgdaInductiveConstructor{refl} suffices. In this case the saved effort is
negligible, but in later results, where it is necessary to use
\AgdaFunction{dual-inv} for rewriting part of the \emph{index} of some type
families, having an implicit rewriting rule allows us to avoid writing some
quite obscure Agda code.

### 2025-12-25 paper/agda.tex
british spelling
global revision

\section{Agda Formalization}
\label{sec:agda}

\input{Type}
\input{Context}
\input{Permutations}
\input{Process}
\input{Semantics}

>>>>

\section{Agda Formalisation}
\label{sec:agda}

In this section we describe the formalisation of \LCP in Agda. We present in
detail only some key parts of the code including the representation of types and
processes, the definition of the operational semantics and the statement of the
main results. The interested reader may find the complete source code in the
public repository.

\input{Type}
\input{Context}
\input{Permutations}
\input{Process}
\input{Congruence}
\input{Reduction}
\input{DeadlockFreedom}
\input{Safety}
\input{Examples}

### 2025-12-25 paper/conclusion.tex
global revision

We have presented a formalisation of \LCP, a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
of \CP~\cite{Wadler14}. Binary sessions can be modeled in \LCP using the
continuation-passing encoding described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
%
Under the assumption that the size of a formalisation is a good measure of its
complexity, the formalisation of \LCP can be qualified as ``simple'' compared to
the existing ones.
%
In this respect, our work casts some doubts on the actual contribution of
context splitting to the complexity of a formalisation and hints instead at
explicit continuations in the underlying model as a simplifying factor.
%
It is quite interesting to note that, among the approaches that have been
proposed to overcome context splitting, the one of \citet{SanoKavanaghPientka23}
makes use of explicit continuations.

We remark that \LCP and the calculus described in the linearity
challenge~\cite{CarboneEtAl24} are fundamentally incomparable. On the one hand,
the challenge only considers a minimal calculus of first-order, monomorphic
sessions while \LCP is a fully-fledged calculus that supports linear, shared,
higher-order, polymorphic channels. On the other hand, the logical foundations
of \LCP prevent the modeling of deadlocks and of sequential processes that own
both endpoints of a session, both of which can be modeled in the calculus of the
challenge.
%
Because of its logical foundations, we like to think of \LCP as a ``canonical''
calculus that deserves its own space in the context of the linearity challenge
alongside with (but not in substitution of) more traditional session calculi.

The simple formalisation of \LCP is a good basis for further investigations. We
have already extended \LCP with support for infinite types and recursive
processes (the formalisation of this extension is publicly available in a
sibling folder of the one for \LCP). In the future, it would be interesting to
formalise the strong normalisation property of \LCP as a consequence of cut
elimination of classical linear logic.

>>>>

We have presented a formalisation of \LCP, a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
of \CP~\cite{Wadler14}. Binary sessions are modeled in \LCP using the
continuation-passing encoding described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
%
Under the assumption that the size of a formalisation is a good measure of its
complexity, the formalisation of \LCP can be qualified as ``simple'' compared to
the existing ones.
%
In this respect, our work casts some doubts on the actual contribution of
context splitting to the complexity of a formalisation and hints instead at
explicit continuations in the underlying model as a simplifying factor.
%
It is quite interesting to note that, among the approaches that have been
proposed to overcome context splitting, the one of \citet{SanoKavanaghPientka23}
makes use of explicit continuations.

We remark that \LCP and the calculus described in the linearity
challenge~\cite{CarboneEtAl24} are incomparable in terms of expressiveness. On
the one hand, the challenge only considers a minimal calculus of first-order,
monomorphic sessions while \LCP is a fully-fledged calculus that supports
linear, shared, higher-order, polymorphic channels. On the other hand, the
logical foundations of \LCP prevent the modeling of deadlocked processes and of
sequential processes that own both endpoints of a session. These processes can
be modeled in the calculus of the challenge, where parallel composition and name
restriction are distinct constructs as in the $\pi$-calculus.
%
Because of its logical foundations, we like to think of \LCP as a ``canonical''
calculus that deserves its own space in the context of the linearity challenge
alongside with (but not in substitution of) more traditional session calculi.

The simple formalisation of \LCP is a good basis for further investigations. We
have already extended \LCP with support for infinite types and recursive
processes (the source code of this extension is publicly available in a sibling
folder of the one for \LCP). In the future, it would be interesting to formalise
the strong normalisation property of \LCP as a consequence of cut elimination of
classical linear logic.

### 2025-12-25 paper/related.tex
global revision

\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10}, a
functional language extended with session communication primitives, along with
an interpreter. \citet{CicconePadovani20} have taken inspiration from Thiemann's
work to formalise a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure depends
on the content of messages.

@@

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalisation of interpreters for
computing models supported by linear type systems. One of the presented case
studies is the formalisation of a fragment of \GV~\citep{Wadler14}. The
separating conjunction we have introduced in \Cref{sec:context-agda} and the
typing of the constructors for the representation of processes in
\Cref{sec:process-agda} have been heavily inspired by this work.

@@

All the formalisations mentioned so far make use of context splitting. In
contrast, \citet{ZalakainDardha21} formalise a generalisation of the linear
$\pi$-calculus which is parametric in a \emph{usage algebra} (to account for
channel sharing/linearity) and that is based on \emph{leftover
typing}~\cite{Allais17}. 
%
With leftover typing a process is typed with respect to an \emph{input context},
which describes the overall available channels, and an \emph{output context},
which describes the residual channels not consumed by the process. In this way,
it is possible to ``concatenate'' typing judgments instead of splitting typing
contexts beforehand. As \citet{ZalakainDardha21} nicely summarise, ``context
splits contain usage information that is already present'' in processes, and
therefore can be omitted. This is true, but only provided that channels are
given a \emph{name} in processes (\citet{ZalakainDardha21} use de Bruijn indices
to this aim). In fact, the co-de Bruijn representation of
processes~\cite{McBride18}, whereby channels are nameless and context splitting
is performed eagerly, can be seen as the ``dual approach'' of leftover typing:
channel names provide information that is already present in their (singleton)
typing context, hence they can be omitted from contexts and processes.

Another work based on context splitting is the one by
\citet{JacobsBalzerKrebbers22}, who formalise a library of \emph{connectivity
graphs} for reasoning and enforcing deadlock freedom in a variant of
\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalisation of
deadlock freedom for a calculus of sessions.

Motivated by the technical difficulties arising from context splits,
\citet{SanoKavanaghPientka23} model a structural version of \CP using an
alternative approach based on \emph{linearity predicates}. The key idea is to
treat typing contexts structurally and to enforce the linear usage of sessions
by checking the syntactical occurrence of channels within processes.
Interestingly, the approach relies on the \emph{explicit naming of
continuations} so as to precisely account for the number of times a channel is
actually used.

\citet{ZackonEtAl25} describe yet another approach to the handling of typing
contexts whereby channels are associated not just with a type but also with a
\emph{tag} (or \emph{multiplicity}) from a given resource algebra for describing
their availability. This approach streamlines context splitting in the sense
that contexts can be treated in an essentially structural way, except for tags
which are conveninently combined using operations from the resource algebra.

@@

\Cref{tab:comparison} summarises the comparison of our formalisation with the
aforementioned ones. The three leftmost colums list the reference papers (sorted
by publication date), the proof assistant and the formalised calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. Some acronyms, like \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
types. It should be noted that the actual calculus being formalised usually
differs from (typically, is a strict subset of) the one identified by the
acronym, but it retains at least some of its key features. We use \SessionPI to
denote (variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
encoded using explicit continuation channels as described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
\SCP for the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
of \CP with explicit continuations.

@@

The rightmost columns of \Cref{tab:comparison} report the \emph{size} (in
kilobytes) of the formalisations as rough (and possibly questionable) estimates
of their complexity. Instead of counting line numbers, which may be affected by
code indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalisation excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalisations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24} when available.
%
Some formalisation define \emph{libraries} which could be reused in different
contexts. In these cases, the size of the library is reported separately next to
the size of the part of the development that uses it.

@@

Based on the available data, we can make the following observations.
%
The use of context splitting, which is very well represented, does not seem to
be a good indicator of the complexity of the formalisation, since it spans the
whole range of sizes. Our formalisation is based on context splitting and it is
the only one that covers \CP in full, yet it is by far the smallest one.

The largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
the ones that adopt a locally nameless representation of channels. In this
formalisations channels are represented in two different ways, depending on
whether they are free or bound. This necessarily entails some duplication of
effort as well as some transformation machinery between the two representations
that contribute to the complexity of the formalisation.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.

>>>>

\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10,Wadler14},
a functional language extended with session communication primitives, along with
an interpreter. \citet{CicconePadovani20} have taken inspiration from this work
to formalise a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure depends
on the content of messages.

@@

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalisation of interpreters for models
supported by linear type systems. One of the presented case studies is the
formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. The
separating conjunction introduced in \Cref{sec:context-agda} and the typing of
the constructors for the representation of processes in \Cref{sec:process-agda}
have been heavily inspired by this work.

@@

\citet{JacobsBalzerKrebbers22} formalise a library of \emph{connectivity graphs}
for reasoning and enforcing deadlock freedom in a variant of
\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalisation of
deadlock freedom for a calculus of sessions.

All the formalisations mentioned so far make use of context splitting. In
contrast, \citet{ZalakainDardha21} formalise a generalisation of the linear
$\pi$-calculus which is parametric in a \emph{usage algebra} (to account for
channel sharing/linearity) and that is based on \emph{leftover
typing}~\cite{Allais17}. 
%
With leftover typing a process is typed with respect to an \emph{input context},
which describes the overall available channels, and an \emph{output context},
which describes the residual channels not consumed by the process. In this way,
it is possible to ``concatenate'' typing judgments instead of splitting typing
contexts beforehand. As \citet{ZalakainDardha21} nicely summarise, ``context
splits contain usage information that is already present in processes,'' and
therefore can be omitted. This is true provided that channels are named
(\citet{ZalakainDardha21} use de Bruijn indices to this aim). In fact, the co-de
Bruijn representation of processes~\cite{McBride18}, whereby channels are
nameless and context splitting is performed eagerly, can be seen as the ``dual
approach'' of leftover typing: channel names provide information that is already
present in their (singleton) typing context, hence they can be omitted from
contexts and processes.

Motivated by the technical difficulties arising from context splits,
\citet{SanoKavanaghPientka23} define a structural version of \CP using an
alternative approach based on \emph{linearity predicates}. The key idea is to
treat typing contexts structurally and to enforce the linear usage of channels
by checking their syntactical occurrence in processes. Interestingly, the
approach relies on the \emph{explicit naming of continuations} so as to
precisely account for the number of times a channel is actually used.

\citet{ZackonEtAl25} describe yet another approach to the handling of typing
contexts whereby channels are associated not just with a type but also with a
\emph{tag} (or \emph{multiplicity}) from a given resource algebra. The tag
summarises the number of allowed usages of a particular channel, including the
possibility that the channel is not available. This approach streamlines context
splitting since contexts can be treated in an essentially structural way, except
for tags which are conveninently combined using operations from the resource
algebra.

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
our own. The three leftmost colums list the reference papers (sorted by
publication date), the proof assistant and the formalised calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. Some acronyms, like \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
types. It should be noted that the actual calculus being formalised usually
differs from (typically, is a strict subset of) the one identified by the
acronym, but it retains at least some of its key features. We use \SessionPI to
denote (variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
encoded using explicit continuation channels as described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
\SCP for the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
of \CP with explicit continuations.

@@

The rightmost columns of \Cref{tab:comparison} report the size (in kilobytes) of
the formalisations as rough (and possibly questionable) estimates of their
complexity. Instead of counting line numbers, which may be affected by code
indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalisation excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalisations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24} when available.
%
Some formalisation define \emph{libraries} which could be reused in different
contexts. In these cases, the size of the library is reported separately next to
the size of the part of the development that uses it.

@@

Based on the available data, we can make the following observations.
%
The use of context splitting, which is very well represented, does not seem to
be a good indicator of the complexity of the formalisation, since it spans the
whole range of sizes. Our formalisation is based on context splitting and it is
the only one that supports all the featurs of \CP, yet it is also the smallest
one.

The largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
the ones that adopt a locally nameless representation of channels. In these
formalisations channels are represented in two different ways, depending on
whether they are free or bound. This necessarily entails some duplication of
effort as well as some transformation machinery between the two representations
that contribute to the complexity of the formalisation.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.
